---
title: "STAT 1261/2260 Homework 4"
author: "Rayan Hassan"
date: "2023-10-03"
output: html_document
---

```{r}
library(babynames)
library(tidyverse)
library(mdsr)
library(Hmisc)
BabynamesDist <- make_babynames_dist()
BabynamesDist
```

# Question 1

Estimations: 
Q1 = 15
Q2 (median) = 31
Q3 = 51

# Question 2

Ashley, Jessica and Lisa seem to have the lower interquartile range for age

# Question 3

The names are ordered by age. For instance, the name Dorothy is last because it is used by the oldest people compared to the other names.

# Question 4

Visual cues used:
- Colors (yellow and red)
- Horizontal bars
- Dots (representing the medians)
- Length (of bars)

# Question 5

```{r}
babynames1 <- babynames %>%
  filter(name == "Taylor") %>%
  group_by(year) %>%
  summarise(
    N = n(),        # N is the number of rows for each name and each year
    total = sum(n), # Note that n is a variable in the babynames data frame
    boys = sum(ifelse(sex == "M", n, 0)) 
  )
babynames1 

new_var <- babynames1 %>%
  mutate(new_var = abs(0.5 - boys/total))
new_var

top_years <- new_var %>%
  arrange(new_var)%>%
  select(year)%>%
  head(10)
top_years



  

```
# Question 6

```{r}
com_male <- BabynamesDist %>%  #com_male stands for "common male"
  filter(sex == "M") %>%
  group_by(name) %>% 
  filter(est_alive_today>0) %>% 
  summarise(
    est_num_alive = sum(est_alive_today),
    q1_age = wtd.quantile(age_today, est_alive_today, probs = 0.25),
    median_age = wtd.quantile(age_today, est_alive_today, probs = 0.5),
    q3_age = wtd.quantile(age_today, est_alive_today, probs = 0.75),

  ) %>%
  filter(est_num_alive>=100000) %>%
  arrange(median_age)%>%
  head(25) # Only keep the top 25


w_plot <- ggplot(
  data = com_male, 
  aes(x = reorder(name, -median_age),  y = median_age)) + 
  xlab(NULL) + 
  ylab("Age (in years)") +
  ggtitle("Youngest Male Names")

w_plot <- w_plot + 
  geom_linerange(
    aes(ymin = q1_age, ymax = q3_age),
    color = "#f3d478", 
    linewidth = 5, 
    alpha = 0.8
    ) +
  coord_flip()

w_plot + 
  geom_linerange(aes(ymin = q1_age, ymax = q3_age),color = "#f3d478", 
                 size = 5, alpha = 0.8) + 
  geom_point(fill = "#ed3324", colour = "white", size = 3, shape = 21) +
  geom_point(aes(y=53,x=24),fill = "#ed3324", colour = "white", size = 3, shape = 21) +
  geom_text(aes(y=58,x=24),label="median",cex=3.8)+
  coord_flip()


```

# Question 7

```{r}

years <- babynames %>%
  group_by(year)%>%
  summarise(total_births=sum(n))


highest_year <- years %>%
  filter(total_births == max(total_births))
highest_year
```

# Question 8 

```{r}
data <- babynames %>%
  filter(year>=1990 & year <=2000, sex=="F")

total_names <- data %>%
  group_by(name)%>%
  summarise(total_births=sum(n))

most_popular <- total_names %>%
  filter(total_births==max(total_births))
most_popular
```

