---
title: "Homework1"
author: "Rayan Hassan"
date: "2023-09-05"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

Exercise 1: Objects 

1) 
```{r}
obj1 <- seq(-3,21,3)
obj2 <- c(1,3,5)
obj3 <- obj2>2.5
obj4 <- "I love pizza."

show(obj1)
show(obj2)
show(obj3)
show(obj4)

length(obj1)
length(obj2)
length(obj3)
length(obj4)
```
2)
```{r,error=TRUE}
obj1 * (-1)
obj1[2:6]
obj1[-2]
obj1 + obj2
obj1 * obj3
obj2 + obj3

sum(obj2)
sum(obj3)

obj1 + obj4


```
Exercise 2: Vectors

```{r}
a <- c(0, 1)
b <- c(TRUE, FALSE)
c <- c("happy", "sad")

a+b
a*b
d <- c(a,b);d
e <- c(a,c);e
g <- c(b,c);g
w <- c(a,b,c); w
```

Exercise 3: Errors


result1: error is the "10.3", it should have been floor(10.3)

result2: error is the "<--", it should have been "<-"

3result: error is the "3" in the beginning, wrong naming of variable

result4: error is not closing the ""

result5: no error but no need to store result in a new variable we can just write data()


Exercise 4: Matrices

1)
```{r}
c <- seq(24,2,-2)
M1 <- matrix(c,3,4)

M2 <- matrix(c,3,4, byrow=TRUE)

show(M1)
show(M2)
```

2)
```{r}
M1[3,1]
```

3)
```{r}
M1[2,]
```

4)
```{r}
M2[,3]
```

Exercise 5: Calculate the following using R


1) 
```{r}
a <- log10(1)
b <- log(1)
a-b
```
2) 
```{r}
x <- sin(pi/3)
y <- cos(pi/3)
z <- x+y
show(z)
```
3) 
```{r}
round(z)
```
4) 
```{r}
round(z,2)
```
5) 
```{r}
2022%%9
```

Exercise 6: Data Frames

1) 
```{r}
df <- data.frame(Name = c("Jacob","James","Jeremy","Jimmy","John"),
                 Year = c(1,2,3,4,3),
                 GPA = c(3.41,3.53,3.32,3.67,4.00),
                 Height = c(1.78,1.85,1.72,1.82,1.68) 
                 )
print(df)
```

2)
```{r}
df$GPA
```

3) 
```{r}
df[3,]
```

4) 
```{r}
df[df$GPA < 3.6,1]
```

5) 
```{r}
s <- df[df$Year==3,3]
s

mean(s)

```

