---
title: "JW Ch. 3 Lab"
author: "Rayan Hassan"
date: "02/01/2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error=TRUE)
```

**Instructions and time-saving hints.** The purpose of this lab is to learn 
linear regression in R.  You will reproduce the commands blocks in
JW Section 3.6 in this Rmarkdown file. **Please do not retype all the R commands
as shown in Section 3.6.**  Instead, go to [link to text for H
3](https://www.statlearning.com/resources-first-edition), find Chapter 3 Lab
text, and cut and paste the text into your Rmarkdown file.  Then break up the R
commands into the R chunks shown in the text.  **Do not put all commands into the
same block!**

After breaking the text into the correct chunks, you may knit the document.
You will notice some problems when knitting.  This is due to the fact that the
pasted commands are meant for the command line, not Rmarkdown.  You must find
Rmarkdown conversions for these problems.  One  problem is the appearance of the
`fix()` commands in the blocks, which displays dataframes in a different window.
As a substitute for `fix()`, please use the `kable` command from library
`knitr`, as in `knitr::kable()`.  The results, however, will be too long to hand
in!  Use this only for your personal use.  For your homework/lab submission,
please instead use `head()` instead of `fix()`.  The TA will be less burdened
that way. 

Since R chunks are independent pieces of code, you will sometimes
produce errors when imitating the format in the book.  For example, the
```abline()``` commands must be grouped in the same chunk as the initial
```plot()``` command.

Congratulations!  When the document is successfully knitted, you should read
Section 3.6 along with the Rmarkdown output.  You should see the commands along
with the output plots.  Together, JW 3.6 and your Rmarkdown file will help you
to learn linear regression in R.  Note that the plotting is done with core R
commands and not with the `ggplot2` package.
Submit your Rmd file along with an unzipped PDF of the result before the deadline.


# Chapter 3 Lab: Linear Regression


```{r}
library(MASS) 
library(ISLR)
```

# Simple Linear Regression

```{r}
#knitr::kable(Boston)
head(Boston)
?Boston
```

```{r} 
# names() lists the names of the columns in Boston
names(Boston)
lm.fit=lm(medv~lstat) # this will give error here because we haven't specified the dataset
```

```{r}
# here we properly fit the model (a linear regression one) where we want to predict house price/median value (medv) based on lstat (percentage of lower-status people)
lm.fit=lm(medv~lstat,data=Boston)
```
```{r}
# using attach here allows us to use medv and lstat without the Boston$ before
attach(Boston)
lm.fit=lm(medv~lstat)
```
```{r}
# here we print the regression model followed by its summary
lm.fit
summary(lm.fit)
```
```{r}
# using names() allows us to see the components of the model. Coef() extracts the intercept and slope. Confint() calculates confidence intervals for coefficients. 
names(lm.fit)
coef(lm.fit)
confint(lm.fit)
```
```{r}
# here we predict the median house value (medv) for lstat=5%, 10%, and 15% respectively. The first way uses a confidence interval and the second way a prediction interval.

# For prediction with confidence interval: 
# we can see for example that for lstat = 5, fit = 29.8 (rounded), which is the predicted median house price. lwr = 29 and upr=30.6 (rounded) which means that we are 95% confident that the mean house price is between $29K and $30.6K

# For prediction with prediction interval:
# we have the same predicted prices. But here, for example for lstat=5 also, we see that lwr = 17.566 and upr = 42.04 (rounded), which means that for an individual house, price can be anywhere between $17.566K and $42.04K, which is a much wider interval. 


predict(lm.fit,data.frame(lstat=(c(5,10,15))), interval="confidence")
predict(lm.fit,data.frame(lstat=(c(5,10,15))), interval="prediction")
```
```{r}
# here we plot lstat vs medv (using scatterplot) and we add a regression line to the plot
plot(lstat,medv)
abline(lm.fit)
abline(lm.fit,lwd=3)
abline(lm.fit,lwd=3,col="red")
```
```{r}
# here we plot lstat vs medv using different customizations in plots
plot(lstat,medv,col="red")
plot(lstat,medv,pch=20)
plot(lstat,medv,pch="+")
plot(1:20,1:20,pch=1:20)
#
```
```{r}
# this sets up a 2x2 grid and plots diagnostic plots
par(mfrow=c(2,2))
plot(lm.fit)
```
```{r}
# plots for residuals vs fitted values and, studendized residuals
plot(predict(lm.fit), residuals(lm.fit))
plot(predict(lm.fit), rstudent(lm.fit))
```
```{r}
# plot for leverage statistics, and the which.max() shows the most influential observation
plot(hatvalues(lm.fit))
which.max(hatvalues(lm.fit))
```

# Multiple Linear Regression

```{r}
# here we are predicting medv based on lstat and age
lm.fit=lm(medv~lstat+age,data=Boston)
summary(lm.fit)
```
```{r}
# here we are predicting medv based on all the variables
lm.fit=lm(medv~.,data=Boston)
summary(lm.fit)
```
```{r}
# vif() is variance intlation factor which tells us how high correlation between predictors is
library(car)
vif(lm.fit)
```
```{r}
# here we remove age variable
lm.fit1=lm(medv~.-age,data=Boston)
summary(lm.fit1)
```
```{r}
# this is just another way to remove age
lm.fit1=update(lm.fit, ~.-age)
```

# Interaction Terms

```{r}
# includes age + lstat + interation (lstat:age)
summary(lm(medv~lstat*age,data=Boston))
```

# Non-linear Transformations of the Predictors

```{r}
# this adds squared term for lstat to model non-linearily
lm.fit2=lm(medv~lstat+I(lstat^2))
summary(lm.fit2)
```
```{r}
# anova compares models
lm.fit=lm(medv~lstat)
anova(lm.fit,lm.fit2)
```
```{r}
#
par(mfrow=c(2,2))
plot(lm.fit2)
```
```{r}
# fits a polynomial regression up to degree 5
lm.fit5=lm(medv~poly(lstat,5))
summary(lm.fit5)
```
```{r}
#
summary(lm(medv~log(rm),data=Boston))
```

# Qualitative Predictors

```{r}
#
fix(Carseats)
names(Carseats)
```
```{r}
# includes all predictors + interaction terms (Income:advertising) and (Price:Age)
lm.fit=lm(Sales~.+Income:Advertising+Price:Age,data=Carseats)
summary(lm.fit)
```
```{r}
# contrasts() just shows how R encodes ShelveLoc numerically
attach(Carseats)
contrasts(ShelveLoc)

```

# Writing Functions

```{r}
# this does not do anything because LoadLibraries is not defined yet
LoadLibraries
LoadLibraries()
```
```{r}
#
LoadLibraries=function(){
 library(ISLR)
 library(MASS)
 print("The libraries have been loaded.")
 }
```

```{r}
# calls the LoadLibraries function and therefore loads the required packages
LoadLibraries
LoadLibraries()
```
