---
title: "JW6 Labs 2 and 3"
author: "Rayan Hassan"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Instructions and time-saving hints.** The purpose of this lab is to learn
linear regression in R.  You will reproduce the commands blocks in JW Section
6.5 in this Rmarkdown file. **Please do not retype all the R commands as shown
in Section 6.5.**  Instead, go to [link to text
](https://www.statlearning.com/resources-first-edition), find Chapter 6 Lab
text, and cut and paste the text into your Rmarkdown file.  Then break up the R
commands into the R chunks shown in the text.  **Do not put all commands into
the same block!**

After breaking the text into the correct chunks, you may knit the document.
You will notice some problems when knitting.  This is due to the fact that the
pasted commands are meant for the command line, not Rmarkdown.  You must find
Rmarkdown conversions for these problems.  One  problem is the appearance of the
`fix()` commands in the blocks, which displays dataframes in a different window.
As a substitute for `fix()`, please use the `kable` command from library
`knitr`, as in `knitr::kable()`.  The results, however, will be too long to hand
in!  Use this only for your personal use.  For your homework/lab submission,
please instead use `head()` instead of `fix()`.  The TA will be less burdened
that way. 

Since R chunks are independent pieces of code, you will sometimes
produce errors when imitating the format in the book.  For example, the
```abline()``` commands must be grouped in the same chunk as the initial
```plot()``` command.

Congratulations!  When the document is successfully knitted, you should read
Section 6.5 along with the Rmarkdown output.  You should see the commands along
with the output plots.  Together, JW 6.5 and your Rmarkdown file will help you
to learn linear regression in R.  Note that the plotting is done with core R
commands and not with the `ggplot2` package. Submit your Rmd file along with an
unzipped PDF of the result before the deadline.


# Chapter 6 Lab 2: Ridge Regression and the Lasso

```{r, include=FALSE}
library(ISLR)
head(Hitters)
names(Hitters)
dim(Hitters)
sum(is.na(Hitters$Salary))
Hitters=na.omit(Hitters)
```

```{r}
x=model.matrix(Salary~.,Hitters)[,-1]
y=Hitters$Salary
```

# Ridge Regression

```{r}
library(glmnet)
grid=10^seq(10,-2,length=100)
ridge.mod=glmnet(x,y,alpha=0,lambda=grid)
```

```{r}
dim(coef(ridge.mod))
```

```{r}
ridge.mod$lambda[50]
coef(ridge.mod)[,50]
sqrt(sum(coef(ridge.mod)[-1,50]^2))
```

```{r}
ridge.mod$lambda[60]
coef(ridge.mod)[,60]
sqrt(sum(coef(ridge.mod)[-1,60]^2))
```

```{r}
predict(ridge.mod,s=50,type="coefficients")[1:20,]
```

```{r}
set.seed(1)
train=sample(1:nrow(x), nrow(x)/2)
test=(-train)
y.test=y[test]
```

```{r}
ridge.mod=glmnet(x[train,],y[train],alpha=0,lambda=grid, thresh=1e-12)
ridge.pred=predict(ridge.mod,s=4,newx=x[test,])
mean((ridge.pred-y.test)^2)
```

```{r}
mean((mean(y[train])-y.test)^2)
ridge.pred=predict(ridge.mod,s=1e10,newx=x[test,])
mean((ridge.pred-y.test)^2)
```


```{r}

ridge.pred=predict(ridge.mod,s=0,newx=x[test,],exact=T,x=x[train,],y=y[train])
mean((ridge.pred-y.test)^2)
lm(y~x, subset=train)
predict(ridge.mod,s=0,exact=T,type="coefficients",x=x[train,],y=y[train])[1:20,]
```


```{r}
set.seed(1)
cv.out=cv.glmnet(x[train,],y[train],alpha=0)
plot(cv.out)
bestlam=cv.out$lambda.min
bestlam
```

```{r}
ridge.pred=predict(ridge.mod,s=bestlam,newx=x[test,])
mean((ridge.pred-y.test)^2)

```
```{r}
out=glmnet(x,y,alpha=0)
predict(out,type="coefficients",s=bestlam)[1:20,]
```

# The Lasso

```{r}
lasso.mod=glmnet(x[train,],y[train],alpha=1,lambda=grid)
plot(lasso.mod)
```

```{r}
set.seed(1)
cv.out=cv.glmnet(x[train,],y[train],alpha=1)
plot(cv.out)
bestlam=cv.out$lambda.min
lasso.pred=predict(lasso.mod,s=bestlam,newx=x[test,])
mean((lasso.pred-y.test)^2)
```

```{r}
out=glmnet(x,y,alpha=1,lambda=grid)
lasso.coef=predict(out,type="coefficients",s=bestlam)[1:20,]
lasso.coef
lasso.coef[lasso.coef!=0]
```

# Chapter 6 Lab 3: PCR and PLS Regression


# Principal Components Regression

```{r}
library(pls)
set.seed(2)
pcr.fit=pcr(Salary~., data=Hitters,scale=TRUE,validation="CV")
```

```{r}
summary(pcr.fit)
```

```{r}
validationplot(pcr.fit,val.type="MSEP")
```

```{r}
set.seed(1)
pcr.fit=pcr(Salary~., data=Hitters,subset=train,scale=TRUE, validation="CV")
validationplot(pcr.fit,val.type="MSEP")
pcr.pred=predict(pcr.fit,x[test,],ncomp=7)
mean((pcr.pred-y.test)^2)

```

```{r}
pcr.fit=pcr(y~x,scale=TRUE,ncomp=7)
summary(pcr.fit)
```

# Partial Least Squares

```{r}
set.seed(1)
pls.fit=plsr(Salary~., data=Hitters,subset=train,scale=TRUE, validation="CV")
summary(pls.fit)
validationplot(pls.fit,val.type="MSEP")
```


```{r}
pls.pred=predict(pls.fit,x[test,],ncomp=2)
mean((pls.pred-y.test)^2)
```


```{r}
pls.fit=plsr(Salary~., data=Hitters,scale=TRUE,ncomp=2)
summary(pls.fit)
```