---
title: "STAT 1261/2260 Homework 6"
author: "Rayan Hassan"
date: "2023-11-05"
output: html_document
---

# Exercise 1

```{r}
library(NHANES)
library(tidyverse)

data1<- NHANES %>%
  select(HardDrugs,Gender,HealthGen,Depressed,SleepHrsNight,AlcoholDay, SmokeNow, Marijuana) %>% na.omit()
glimpse(data1)
```
# Exercise 1


```{r}
set.seed(99)
train1 <- data1 %>% sample_frac(size = 0.8)
test1 <- data1 %>% setdiff(train1)


library(glmnet)
form<- as.formula(
  "HardDrugs ~Gender+HealthGen+Depressed+SleepHrsNight+AlcoholDay+SmokeNow+Marijuana")
predictors <- model.matrix(form, data = train1) 
cv.fit <- cv.glmnet(predictors, train1$HardDrugs, family = "binomial", type = "class")
cv.fit$lambda.1se

plot(cv.fit)


lambda_opt=cv.fit$lambda.1se
mod_lr <- glmnet(predictors, train1$HardDrugs, family = "binomial", lambda = lambda_opt)
mod_lr$beta

form2 <- as.formula(HardDrugs~Gender+Depressed+AlcoholDay+Marijuana)
mod_lr2 <- glm(form2, data=train1,family=binomial)
summary(mod_lr2)
```
# Exercise 2

The variable that is most significant is MarijuanaYes with the smallest p-value of < 2e-16

# Exercise 3

The coefficient of MarijuanaYes is 3.11054 (positive). So it is a positive direction which indicated that individuals who use Marijuana are more likely to use hard drugs. In terms of strength, MarijuanaYes has a very small p-value of less then 2e-16 (almost close to zero), which suggests that the relationship between Marijuana and HardDrugs is extremely strong.

# Exercise 4

Gendermale has a very small p-value (around 6.5e-5) which indicates that Gendermale is a significant variable. Also, it has a coefficient of 0.51475, which is positive. This suggests that males are more likely to use hard drugs and that the relationship between both is very strong

```{r}
library(rpart)
library(rpart.plot)

mod_tree <- rpart(form2,data=train1)
rpart.plot(mod_tree)
```

# Exercise 5

The most important predictor seems to be Marijuana. Only 4% of people that don't do Marijuana do Hard Drugs versus 46% that smoke Marijuana. This difference is huge and shows how much smoking Marijuana leads to hard drugs.  

# Exercise 6

The two subgroups that are most likely to use hard drugs are Marijuana which we explained before (so if person smokes marijuana) and if person is Depressed. 58% of people that do Marijuana and that are Depressed do Hard Drugs. Whereas 41% people who do Marijuana but are not Depressed do Hard Drugs. Which shows that in addition to smoking Marijuana, Depression is also a significant factor. 

# Exercise 7

```{r}

data2<- NHANES %>%
  select(Gender,Age,Race1,Education,MaritalStatus,HHIncomeMid,Poverty,HomeOwn,Weight,Height,
         BMI,Pulse,BPSysAve,BPDiaAve,Diabetes,HealthGen,DaysPhysHlthBad,DaysMentHlthBad,
         Depressed,SleepHrsNight,SleepTrouble,AlcoholDay,Smoke100,Marijuana,HardDrugs) %>% 
  drop_na()

train2 <- data2 %>% sample_frac(size = 0.8, fac=HardDrugs)
test2 <- data2 %>% setdiff(train2)

form_full<- as.formula(HardDrugs~Gender+Age+Race1+Education+MaritalStatus+HHIncomeMid+Poverty+HomeOwn+Weight+Height+BMI+Pulse+BPSysAve+BPDiaAve+Diabetes+HealthGen+DaysPhysHlthBad+DaysMentHlthBad+Depressed+SleepHrsNight+SleepTrouble+AlcoholDay+Smoke100+Marijuana)
predictors <- model.matrix(form_full, data = train2) 
cv.fit <- cv.glmnet(predictors, train2$HardDrugs, family = "binomial", type = "class")
cv.fit$lambda.1se

plot(cv.fit)

lambda_opt=cv.fit$lambda.1se
mod_lr2 <- glmnet(predictors, train2$HardDrugs, family = "binomial", lambda = lambda_opt)
mod_lr2$beta

```

# Exercise 7

```{r}
mod_tree2 <- rpart(form_full,data=train2)
rpart.plot(mod_tree2)
```

# Exercise 8

```{r}
confusion_matrix <- function(data,y,mod){
  confusion_matrix <- data %>% 
  mutate(pred = predict(mod, newdata = data, type = "class"),
         y=y) %>%
  select(y,pred) %>% table()
}

misclass <- function(confusion){
  misclass <- 1- sum(diag(confusion))/sum(confusion)
  return(misclass)
}


c <- confusion_matrix(test2, test2$HardDrugs, mod_tree2)
c


misclass(c)

```









